Social Participation Ontology: community documentation, enhancements and use examples

### 2015-01-09 paper.tex
feita toda a exposicao da ops

\usepackage{graphicx}% Include figure files
\usepackage{grffile}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines
\usepackage{multirow}
\usepackage{color} % for the notes
\usepackage{etex}
\usepackage{hyperref}
\reserveinserts{58}
\usepackage{morefloats}
\usepackage[section]{placeins}

@@

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified
\newcommand{\ops}{\textsc{ops}}
\newcommand{\vcps}{\textsc{vcps}}
\newcommand{\owl}{\textsc{owl}}
\newcommand{\sparql}{\textsc{s}par\textsc{ql}}
\newcommand{\bfo}{\textsc{bfo}}
\newcommand{\dbpedia}{\textsc{db}pedia}
\newcommand{\foaf}{\textsc{foaf}}
\newcommand{\ict}{\textsc{ict}}
\newcommand{\html}{\textsc{html}}
\newcommand{\node}{\textsc{n}ode.js}
\newcommand{\facebook}{\textsc{f}acebook}
\newcommand{\twitter}{\textsc{t}witter}
\newcommand{\wwwc}{\textsc{w3c}}
\newcommand{\skos}{\textsc{skos}}
\newcommand{\etherpad}{\textsc{e}therpad}
\newcommand{\ogp}{\textsc{ogp}}
\begin{abstract}
    Participatory democracy advances in virtually all governments. South America presents a prominent context with mixed culture and social predisposition. At 2012, civil, academic and governmental parties started elaborating the ``Common Vocabulary of Social Participation'' (\vcps\ from the Brazilian name \emph{Vocabul\'ario Comum de Participa\c{c}\~ao Social}), as a public and online process. By May, 2013, first reference documents were publicized, together with a preliminary \owl\ code,
logos, and a diagram of a general ``public consultation''.
The Corais Platform kept online records of the process, like discussions and preparation of texts. 
This article exposes this material and proposes elementary unfolding, highlighting the ``Social Participation Ontology'' (\ops\ from the Brazilian name \emph{Ontologia de Participa\c{c}\~ao Social}). To exhibit this new ontology, this steps were considered: completion of \vcps\ \owl\ code for preliminary \ops, with necessary translations and corrections; correction of ontological contradictions and enhancements of class names and labels in Portuguese, Spanish and English; a toy expansion of the ontology by further specifying classes; use cases of \ops\ by researchers and public managers; and use examples regarding knowledge organization and a \sparql\ endpoint. Ongoing work involves incorporation by the official Brazilian social participation federal portal, incorporation by civil participative organizations, linkage to other participative ontologies. \ops\ is being used as an upper ontology, and all classes linked to \foaf\ and \bfo\ as higher upper ontologies.
\end{abstract}

@@

\begin{itemize}
    \item The usage of the word ``vocabulary'' can lead to confusion is some situations as \ops\ is an \owl\ ontology (not, for example, a \skos\ vocabulary).
    \item Documentation seems inconsistent when an ontology is repeatedly called a vocabulary.
    \item \ops\ is, in fact, an ontology, with a vocabulary, a taxonomic organization and properties further relating the terms.
    \item This coherent naming is a prerequisite for for academic acceptance and further formal adoption by government instances, such as the Brazilian Federal Portal of Open Data~\cite{dadosGov} and of Social Participation~\cite{participabr}.
\end{itemize}

@@

This section makes four considerations about the \ops: first is label standardization and an exhibition of implemented and unimplemented \vcps\ classes, properties and restrictions. Features present in Figure~\ref{fig:v1}, but not present in the community release, are fully described in Section~\ref{impl}. Elementary further specification of subclasses is addressed in Section~\ref{downwards}. Examples of usage are addressed in Section~\ref{examplesUsage}.

For completeness of exposition and deference to CVSP community efforts, consideration of upper ontologies is at Section~\ref{upper} and the public consultation model depicted in Figure~\ref{fig:consult} is considered at Section~\ref{public}.

\subsection{Standardization and exhibition of implemented features}\label{impl}
Without explicit criteria, CVSP IRI was \url{http://lumii.lv/ontologies/Corais.owl}. OPS IRI was chosen to be \url{http://purl.org/socialparticipation/ops} for the following reasons:
\begin{itemize}
    \item This IRI is directly related to the ontology name (OPS).
    \item This IRI, also an URL, is independent from government and other political parties. This is important to coalesce with interested parties: the Brazilian Federal Social Participation Portal\cite{participa}, Brazilian government repository of vocabularies and ontologies~\cite{vocab}, civil society and NGOs.
    \item The IRI, when reached via HTTP, can be redirected to where current documentation is held.
\end{itemize}

Labels in the languages of interest should be written in label fields. Even so, class names can be friendly to (potential) users, bearing the attention not to take the class name as the label or as a meaning restriction.

For standardization, all classes were written in CamelCase and without accents. Class names were written in English to ease internationalization, adoption and maturation. Labels were written in Portuguese, English and Spanish. Therefore, class names changed, receiving respective labels and a plain textual short explanation.
Table~\ref{ospClasses} exhibits all classes is OSP.

\begin{table*}
    \footnotesize
  \centering
  \caption{Classes of the OSP (Ontology of Social Participation). These are core concepts in the ontology. Along with the taxonomic structure exposed in Figure~\ref{}, these classes are related by the properties in Table~\ref{ospProps}.}
  \begin{tabular}{|l|l||p{2.2cm}|p{2.2cm}|p{1.8cm}||p{4cm}|}\hline
{\bf OSP class name} & {\bf CVSP class name} & {\bf pt-br label} & {\bf es label} & {\bf en label} & {\bf definition} \\\hline\hline
Problem & Problema & Problema & Problema & Problem & the problem that the Action aims to solve \\\hline
Result & Resultados & Resultado & Resultado & Result & the result obtained with action realization \\\hline
Theme & Tema & Tema & Tema & Theme & the theme in focus by action \\\hline
Cause & Causa & Causa & Causa &  Cause & the motivation for the execution of the action \\\hline
Action & Acao & A\c{c}\~ao & Acci\'on & Action & what is done in terms os social participation \\\hline
Solution & Solucao & Solução & Soluci\'on & Solution & solucion achieved with the action \\\hline
ParticipationCharacteristic & NivelDeParticipacao & Caracter\'istica de Participa\c{c}\~ao & Caracter\'istica de Participaci\'on & Participation Characteristic & the way the participation of s specific actor, is happening \\\hline
ActionScope & Espa\c{c}o de A\c{c}\~ao & Escopo de A\c{c}\~ao & Ambito de Accio\'on & Action Scope & the scope os Action execution \\\hline
Role & Papel & Papel & Rol & Role & the role of the actor \\ \hline
Supporter & Apoiador & Apoiador & Apoyador & Supporter & supports cause with resources of any kind (e.g. cognitive, financial, equipments) \\ \hline
Initiator & Iniciador & Iniciador & Iniciador & Initiator & originates cause, individually or collaborativelly \\ \hline
Executor & Executor & Executor & Ejecutor & Executor & performs action directly and is responsible for results \\ \hline
SocialActor & Ator & Ator Social & Actor Social & Social Actor & entity that might have a role \\ \hline
Organization & Organizacao & Organiza\c{c}\~ao & Organizaci\'on & Organization & social actor is a group of individuals, organized formally or informally (e.g. movements, collectives) \\ \hline
Person & Pessoa & Pessoa & Persona & Person & a person (social actor is a person) \\ \hline
  \end{tabular}
  \label{ospClasses}
\end{table*}

OSP Properties fit headlessCamelCase format, are readable in English (to ease internationalization, adoption and maturation), and have well defined domains and ranges. Table~\ref{ospProps} is dedicated to OSP properties, with labels in English, Portuguese and Spanish.

@@

All restrictions in OSP are existential and all properties yield restrictions, except hasParticipationCharacteristic. Table~\ref{ospRestr} is dedicated to OSP restrictions.

\begin{table}
  \centering
  \caption{Restrictions of the OSP (Ontology of Social Participation). All OSP restrictions are existential (\texttt{owl:someValuesFrom)}.}
  \begin{tabular}{|l|l|l|}\hline
{\bf subject} & {\bf predicate} & {\bf object} \\\hline\hline
Initiator    & startsCause        & Cause\\\hline
Supporter    & supportsCause      & Cause\\\hline
Executer     & executesAction     & Action\\\hline
Solution     & solvesProblem      & Problem\\\hline
SocialActor  & hasRole            & Role\\\hline
Action       & producesResults    & Results\\\hline
Result       & composesSolution   & Solution\\\hline
Cause        & dasAction          & Action\\\hline
Action       & belongsToScope     & Scope\\\hline
Cause        & hasTheme           & Theme\\\hline
Cause        & proposesSolution   & Solution\\\hline
Problem      & generatesCause     & Cause\\\hline
  \end{tabular}
  \label{ospRestr}
\end{table}

A comparison of the CVSP OWL code~\cite{owlCCPtg} 
with the diagram in Figure~\ref{fig:v1}, which reflects official CVSP documentation, revealed that a class, two properties and three restrictions were not implemented. These were fully implemented in OPS. These are the missing classes, properties and restrictions missing in CVSP and that were implemented in OSP:

\begin{itemize}
    \item Class: ParticipationCharacteristic.
    \item Property: hasRole.
    \item Property: composesSolution.
    \item Restriction: Role hasParticipationCharacteristic some ParticipationCharacteristic.
    \item Restriction: Results composesSolution some Solution.
    \item Restriction: Problem generatesCause some Cause.
\end{itemize}

Also, in CVSP, existential restrictions were  written as ``min 1''. 
These were changed to the standard ``some'' existential restriction.

OSP is available online~\cite{owlOSP}. To ease navigation of the ontology by interested parties, it is also available in the Webprotege interface~\cite{owlOSPwp}. The diagram of OSP taxonomic structure is exposed in Figure~\ref{fig:owlCC}.

Examples of upper ontologies usage with OSP is under development and should be available in a dedicated work, as possibilities should be inspected carefully. Pertinent to OSP, but not upper ontologies, are FOAF~\cite{foaf} (for describing people and things they do) and Dublin Core~\cite{dublin} (for describing resources, with special attention to digital resources).

\subsection{OSP growth}\label{downwards}

@@

\begin{table*}
  \centering
  \caption{New classes considered for expansion of the OSP. The taxonomic organization of these classes, within OSP can be oberved in Figure~\ref{fig:owlExp}. Further information is in Section~\ref{downwards}.}
  \begin{tabular}{|l|l|p{5cm}|p{5cm}|}\hline
{\bf new class} & {\bf subclass of} & {\bf description} & {\bf further notes} \\\hline\hline
SocialNetwork & Organization & a social structure made up of social actors (such as individuals or organizations) and a set of dyadic ties between these actors & --//--\\\hline
FreeScaleNetwork & SocialNetwork & a network whoose connectivity follows a power law & disjoint with UniformrandomNetwork and GeographicNetwork \\ \hline
UniformRandomNetwork & SocialNetwork & also known as Erd\"os Renyi network, this network sets, with equal propability, an edge between each pair of nodes & disjoint with FreeScaleNetwork and GeographicNetwork\\\hline
GeographicNetwork & SocialNetwork & a network whose connectivity is related to the distance of nodes in a metric space & Disjoint with both FreeScaleNetwork and UniformRandomNetwork \\\hline
SmallWorldNetwork & SocialNetwork & a network where most nodes can be reached from other nodes with few hops or steps & not disjoint with FreeScaleNetwork, UniformNetwork and GeographicNetwork\\\hline\hline
InformalOrganization & Organization & an organization that is not formalized & disjoint with Intitution \\ \hline
Mob & InformalOrganization & a crowd of individuals & --//-- \\\hline
GiantMob & Mob & a crowd with more than 10,000 individuals & --//--\\\hline
DownloadedMob & Mob & a Mob convoqued by a network & this is a defined class, by being a network and being the subject of the relation convoquedBy with object Network \\ \hline\hline
Institution & Organization & a mechanism of social order that governs a set of individuals & disjoint with InformalOrganization \\\hline
PublicInstitution & Institution & an institution backed through public funds and controlled by the state & disjoint with Private Institution \\ \hline
PrivateInstitution & Institution & an institution backed through private fundings and controlled by private parties & disjoint with PublicInstitution \\ \hline
AcademicInstitution & Institution & an institution dedicated to education and research, which grants academic degrees & --//-- \\ \hline
NGO & Institution & a legally consituted corporation created by natural or legal people that operate independently from anu form of government & --//-- \\\hline
SpuriousInstitution & Institution & an institution that holds prominent illegitimate or corrupt characteristcs & --//-- \\\hline
ExoticInstitution & Institution & and institution that does not fir previous classes or is characterized by a very unique traces & --//-- \\\hline\hline
VoluntaryExecutor & Executor & an executor that receives no formal reward for the tasks & disjoint with PaidExecutor \\ \hline
PaidExecutor & Executor & an Executor that receives formal reward for the tasks accomplished & a defined class, being an Executor and the subjecto of a receivesFrom predicate with SocialActor as object\\ \hline
  \end{tabular}
  \label{ospFooClass}
\end{table*}
%\subsubsection{SparQL examples of usage}

@@

Imagined examples are useful to make these points clear:
\begin{itemize}
    \item Suppose a public SparQL endpoint unifies several participation instances by means of OSP. Thus, the number of participants (Social Actors) is publicly available. Also, depending on the platforms, one can observe how many are individuals, how many are organizations, and understand to which extent the corporative influence is explicit. One can observe how many of the participants are the same in each platform, and what roles they take, and make conclusions about how much society is really participating or if these processes are manipulated by the same agents. One can also gaze uppon the problems being discussed and which solutions are being proposed, therefore easing the sense of what is being considered important and valid as public discussions. This list of possibilities is endless, specially when OSP variations and expansions are considered.
    \item Suppose a person, say Jessica, has a new proposal for a participative system that uses OSP. She can have a quite concise understanding of the concepts involved, and how they relate, by means of OSP. She can now make very objective observations and deliver clear suggestions that relate directly to the system being used. She can even make an OSP variation or another ontology, as a way to confront the schemes.
    \item Supose there is an automatic system for exhibiting indicators about social participation (how effective it has been, how wide the scope of interests, etc). Instances that are integrated by OSP are queried for information and, for example, this system registers any organization involved as a social actor. In the case of the expanded OSP exposed above, for example, the system registers any mob, whose incidence was recorded in the database as a \texttt{DownloadedMob}, as related to some social network, be the network known or not.
\end{itemize}

@@

>>>>

\usepackage{graphicx}% Include figure files
\usepackage{grffile}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines
\usepackage{multirow}
\usepackage{color} % for the notes
\usepackage{etex}
\usepackage{hyperref}
\reserveinserts{58}
\usepackage{morefloats}
\usepackage[section]{placeins}
\usepackage{xcolor}
\hypersetup{
        colorlinks,
            linkcolor={red!50!black},
                citecolor={blue!50!black},
                    urlcolor={blue!80!black}
                }

@@

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified
\newcommand{\ops}{\textsc{ops}}
\newcommand{\opsi}{O\textsc{ps}}
\newcommand{\vcps}{\textsc{vcps}}
\newcommand{\owl}{\textsc{owl}}
\newcommand{\sparql}{\textsc{s}par\textsc{ql}}
\newcommand{\bfo}{\textsc{bfo}}
\newcommand{\dbpedia}{\textsc{db}pedia}
\newcommand{\foaf}{\textsc{foaf}}
\newcommand{\ict}{\textsc{ict}}
\newcommand{\html}{\textsc{html}}
\newcommand{\node}{\textsc{n}ode.js}
\newcommand{\facebook}{\textsc{f}acebook}
\newcommand{\twitter}{\textsc{t}witter}
\newcommand{\wwwc}{\textsc{w3c}}
\newcommand{\skos}{\textsc{skos}}
\newcommand{\etherpad}{\textsc{e}therpad}
\newcommand{\ogp}{\textsc{ogp}}
\newcommand{\iri}{\textsc{iri}}
\newcommand{\uri}{\textsc{uri}}
\newcommand{\urll}{\textsc{url}}
\newcommand{\ngo}{\textsc{ngo}}
\newcommand{\http}{\textsc{http}}
%\newcommand{\t}[1]{\texttt{#1}}
\begin{abstract}
    Participatory democracy advances in virtually all governments. South America presents a prominent context with mixed culture and social predisposition. At 2012, civil, academic and governmental parties started elaborating the ``Common Vocabulary of Social Participation'' (\vcps\ from the Brazilian name \emph{Vocabul\'ario Comum de Participa\c{c}\~ao Social}), as a public and online process. By May, 2013, first reference documents were publicized, together with a preliminary \owl\ code,
logos, and a diagram of a general ``public consultation''.
The Corais Platform kept online records of the process, like discussions and preparation of texts. 
This article exposes this material and proposes elementary unfolding, highlighting the ``Social Participation Ontology'' (\ops\ from the Brazilian name \emph{Ontologia de Participa\c{c}\~ao Social}). To exhibit this new ontology, this steps were considered: completion of \vcps\ \owl\ code for preliminary \ops, with necessary translations and corrections; correction of ontological contradictions and enhancements of class names and labels in Portuguese, Spanish and English; a toy expansion of the ontology by further specifying classes; use cases of \ops\ by researchers and public managers; and use examples regarding knowledge organization and a \sparql\ endpoint. Ongoing work involves incorporation by the official Brazilian social participation federal portal, incorporation by civil participative organizations, linkage to other participative ontologies. \ops\ is being used as an upper ontology, and all classes linked to \foaf\ and \bfo\ as higher upper ontologies.
\end{abstract}

@@

\begin{itemize}
    \item The usage of the word ``vocabulary'' can lead to confusion is some situations as \ops\ is an \owl\ ontology (not, for example, a \skos\ vocabulary).
    \item Documentation seems inconsistent when an ontology is repeatedly called a vocabulary.
    \item \opsi\ is, in fact, an ontology, with a vocabulary, a taxonomic organization and properties further relating the terms.
    \item This coherent naming is a prerequisite for for academic acceptance and further formal adoption by government instances, such as the Brazilian Federal Portal of Open Data~\cite{dadosGov} and of Social Participation~\cite{participabr}.
\end{itemize}

@@

This section makes considerations about \ops\ label standardization and implemented classes, properties and restrictions. Features present in Figure~\ref{fig:v1}, but not present in \vcps, are fully described in Section~\ref{impl}. Examples of usage are addressed in Section~\ref{ospUtil}.

For completeness of exposition and deference to \vcps\ community efforts, consideration of upper ontologies is at Appendix~\ref{upper} and the public consultation model depicted in Figure~\ref{fig:consult} is considered in Appendix~\ref{ap:public}.

\subsection{Standardization and exhibition of implemented features}\label{impl}
Without explicit criteria, \vcps\ \uri\ was \url{http://lumii.lv/ontologies/Corais.owl}. \opsi\ \uri\ was chosen to be \url{http://purl.org/socialparticipation/ops} for the following reasons:
\begin{itemize}
    \item This \uri\ is directly related to the ontology name (\ops).
    \item This \uri, also an \urll, is independent from government and other political associations. This is important to coalesce with interested parties: the Brazilian Federal Social Participation Portal\cite{participa}, Brazilian government repository of vocabularies and ontologies~\cite{vocab}, academic groups,  \ngo s, and non-organized civil society.
    \item This \uri, when reached via \http, can be redirected to where current documentation is held, as it is hosted by \url{http://purl.org}.
\end{itemize}

Labels in the languages of interest should be written in label fields. Even so, \ops\ class names should be friendly to users, bearing the attention not to take the class name as the label or as a meaning restriction.

For standardization, all classes are written in CamelCase~\cite{cc} in plain English to ease internationalization, adoption and maturation. Labels are written in Portuguese, English and Spanish. Therefore, class names changed (for their \uri s), receiving respective labels (\texttt{rdfs:label}) in each language and a textual short explanation (\texttt{rdfs:comment}) in English. 
Table~\ref{ospClasses} exhibits all classes is \ops.

\opsi\ property names fit headlessCamelCase~\cite{hcc} format, readable in English (to ease internationalization, adoption and maturation), and some of them have defined domains and ranges. Table~\ref{ospProps} is dedicated to \ops\ properties, with labels in English, Portuguese and Spanish.

In the first versions of \ops, all properties yielded existential restrictions, except hasParticipationCharacteristic. Although such efforts were aimed at handling a better defined \ops, further developments and discussions revealed that these restrictions made \ops\ rigid, a bit more complicated, and was not of much help, at least in this stage of \ops development and adoption. The result is that all restrictions were removed. Appendix~\ref{ap:restr} and Table~\ref{ospRestr} exhibit all dropped \ops\ restrictions.

\begin{table*}
    \footnotesize
  \centering
  \caption{Classes of the OSP (Ontology of Social Participation). These are core concepts in the ontology. Along with the taxonomic structure exposed in Figure~\ref{}, these classes are related by the properties in Table~\ref{ospProps}.}
  \begin{tabular}{|p{1.8cm}|p{1.6cm}||p{2.2cm}|p{2.2cm}|p{1.8cm}||p{4cm}|}\hline
{\bf OSP class name} & {\bf CVSP class name} & {\bf pt-br label} & {\bf es label} & {\bf en label} & {\bf definition} \\\hline\hline
Problem & Problema & Problema & Problema & Problem & the problem that the Action aims to solve \\\hline
Result & Resultados & Resultado & Resultado & Result & the result obtained with action realization \\\hline
Theme & Tema & Tema & Tema & Theme & the theme in focus by action \\\hline
Cause & Causa & Causa & Causa &  Cause & the motivation for the execution of the action \\\hline
Action & Acao & A\c{c}\~ao & Acci\'on & Action & what is done in terms os social participation \\\hline
Solution & Solucao & Solução & Soluci\'on & Solution & solucion achieved with the action \\\hline
Participation-Characteristic & NivelDe-Participacao & Caracter\'istica de Participa\c{c}\~ao & Caracter\'istica de Participaci\'on & Participation Characteristic & the way the participation of s specific actor, is happening \\\hline
ActionScope & Espa\c{c}o de A\c{c}\~ao & Escopo de A\c{c}\~ao & Ambito de Accio\'on & Action Scope & the scope os Action execution \\\hline
Role & Papel & Papel & Rol & Role & the role of the actor \\ \hline
Supporter & Apoiador & Apoiador & Apoyador & Supporter & supports cause with resources of any kind (e.g. cognitive, financial, equipments) \\ \hline
Initiator & Iniciador & Iniciador & Iniciador & Initiator & originates cause, individually or collaborativelly \\ \hline
Executor & Executor & Executor & Ejecutor & Executor & performs action directly and is responsible for results \\ \hline
SocialActor & Ator & Ator Social & Actor Social & Social Actor & entity that might have a role \\ \hline
Organization & Organizacao & Organiza\c{c}\~ao & Organizaci\'on & Organization & social actor is a group of individuals, organized formally or informally (e.g. movements, collectives) \\ \hline
Person & Pessoa & Pessoa & Persona & Person & a person (social actor is a person) \\ \hline
  \end{tabular}
  \label{ospClasses}
\end{table*}

@@

A comparison of the \vcps\ \owl\ code~\cite{owlCCPtg} 
with the diagram in Figure~\ref{fig:diaorig}, which reflects official \vcps\ documentation, revealed that a class, two properties and three restrictions were not implemented. These were fully implemented in \ops, though the restrictions were removed. These are the missing classes, properties (restrictions missing in \vcps\ and implemented in preliminary \ops\ versions are exposed in Appendix~\ref{ap:restr}):

\begin{itemize}
    \item Class: {\tt    ops:ParticipationCharacteristic}.
    \item Property: {\tt ops:hasRole}.
    \item Property: {\tt ops:composesSolution}.
\end{itemize}

\opsi\ is available online~\cite{owlOSP}. To ease navigation of the ontology by interested parties, it is also available in the Webprotege interface~\cite{owlOSPwp}. The diagram of \ops\ taxonomic structure is exposed in Figure~\ref{fig:owlCC}.

Upper ontologies usage with \ops\ is under development and should be available in a dedicated work, as possibilities should be inspected carefully. Pertinent and already used as upper ontologies for \ops are \foaf~\cite{foaf} (for linking and describing people and things they do) and \bfo~\cite{bfo} (``designed for use in supporting information retrieval, analysis and integration in scientific and other domains'' as stated on their documentation). Properties were not related to upper ontologies as reasonable relations are still being searched for. Table~\ref{ospClasses} exposes relations of classes to upper ontologies.

Figure~\ref{fig:v1} is a complete diagram of current \ops: classes, properties and relations to \foaf\ and \bfo. Actually, Figure~\ref{fig:v1} is more informative than \ops\ \owl\ code, as restrictions were removed and not all properties have defined domain and range. Therefore, the diagram is the remaining source of relations envisioned by \ops\ creators.

@@

%\subsubsection{SparQL examples of usage}

@@

Imagined examples are useful to make these points clear:
\begin{itemize}
    \item Suppose a public SparQL endpoint unifies several participation instances by means of OSP. Thus, the number of participants (Social Actors) is publicly available. Also, depending on the platforms, one can observe how many are individuals, how many are organizations, and understand to which extent the corporative influence is explicit. One can observe how many of the participants are the same in each platform, and what roles they take, and make conclusions about how much society is really participating or if these processes are manipulated by the same agents. One can also gaze uppon the problems being discussed and which solutions are being proposed, therefore easing the sense of what is being considered important and valid as public discussions. This list of possibilities is endless, specially when OSP variations and expansions are considered.
    \item Suppose a person, say Jessica, has a new proposal for a participative system that uses OSP. She can have a quite concise understanding of the concepts involved, and how they relate, by means of OSP. She can now make very objective observations and deliver clear suggestions that relate directly to the system being used. She can even make an OSP variation or another ontology, as a way to confront the schemes.
    \item Supose there is an automatic system for exhibiting indicators about social participation (how effective it has been, how wide the scope of interests, etc). Instances that are integrated by OSP are queried for information and, for example, this system registers any organization involved as a social actor. In the case of the expanded OSP exposed above, for example, the system registers any mob, whose incidence was recorded in the database as a \texttt{DownloadedMob}, as related to some social network, be the network known or not.
\end{itemize}
\subsection{OSP growth}\label{downwards}

\begin{table*}
  \centering
  \caption{New classes considered for expansion of the OSP. The taxonomic organization of these classes, within OSP can be oberved in Figure~\ref{fig:owlExp}. Further information is in Section~\ref{downwards}.}
  \begin{tabular}{|l|l|p{5cm}|p{5cm}|}\hline
{\bf new class} & {\bf subclass of} & {\bf description} & {\bf further notes} \\\hline\hline
SocialNetwork & Organization & a social structure made up of social actors (such as individuals or organizations) and a set of dyadic ties between these actors & --//--\\\hline
FreeScaleNetwork & SocialNetwork & a network whoose connectivity follows a power law & disjoint with UniformrandomNetwork and GeographicNetwork \\ \hline
UniformRandomNetwork & SocialNetwork & also known as Erd\"os Renyi network, this network sets, with equal propability, an edge between each pair of nodes & disjoint with FreeScaleNetwork and GeographicNetwork\\\hline
GeographicNetwork & SocialNetwork & a network whose connectivity is related to the distance of nodes in a metric space & Disjoint with both FreeScaleNetwork and UniformRandomNetwork \\\hline
SmallWorldNetwork & SocialNetwork & a network where most nodes can be reached from other nodes with few hops or steps & not disjoint with FreeScaleNetwork, UniformNetwork and GeographicNetwork\\\hline\hline
InformalOrganization & Organization & an organization that is not formalized & disjoint with Intitution \\ \hline
Mob & InformalOrganization & a crowd of individuals & --//-- \\\hline
GiantMob & Mob & a crowd with more than 10,000 individuals & --//--\\\hline
DownloadedMob & Mob & a Mob convoqued by a network & this is a defined class, by being a network and being the subject of the relation convoquedBy with object Network \\ \hline\hline
Institution & Organization & a mechanism of social order that governs a set of individuals & disjoint with InformalOrganization \\\hline
PublicInstitution & Institution & an institution backed through public funds and controlled by the state & disjoint with Private Institution \\ \hline
PrivateInstitution & Institution & an institution backed through private fundings and controlled by private parties & disjoint with PublicInstitution \\ \hline
AcademicInstitution & Institution & an institution dedicated to education and research, which grants academic degrees & --//-- \\ \hline
NGO & Institution & a legally consituted corporation created by natural or legal people that operate independently from anu form of government & --//-- \\\hline
SpuriousInstitution & Institution & an institution that holds prominent illegitimate or corrupt characteristcs & --//-- \\\hline
ExoticInstitution & Institution & and institution that does not fir previous classes or is characterized by a very unique traces & --//-- \\\hline\hline
VoluntaryExecutor & Executor & an executor that receives no formal reward for the tasks & disjoint with PaidExecutor \\ \hline
PaidExecutor & Executor & an Executor that receives formal reward for the tasks accomplished & a defined class, being an Executor and the subjecto of a receivesFrom predicate with SocialActor as object\\ \hline
  \end{tabular}
  \label{ospFooClass}
\end{table*}

@@

\section{Restrictions in \vcps\ and initial \ops\ which were removed for the time being}\label{ap:restr}

These restrictions were part of \vcps\ documentation, but not implemented in \vcps\ \owl\ code. These were implemented in preliminary \ops\ implementation, although removed from current \ops\ as explained in Section~\ref{impl}:
\begin{itemize}
    \item Restriction: Role hasParticipationCharacteristic some ParticipationCharacteristic.
    \item Restriction: Results composesSolution some Solution.
    \item Restriction: Problem generatesCause some Cause.
\end{itemize}
\begin{table}
  \centering
  \caption{Restrictions of the OSP (Ontology of Social Participation). All OSP restrictions are existential (\texttt{owl:someValuesFrom)}.}
  \begin{tabular}{|l|l|l|}\hline
{\bf subject} & {\bf predicate} & {\bf object} \\\hline\hline
Initiator    & startsCause        & Cause\\\hline
Supporter    & supportsCause      & Cause\\\hline
Executer     & executesAction     & Action\\\hline
Solution     & solvesProblem      & Problem\\\hline
SocialActor  & hasRole            & Role\\\hline
Action       & producesResults    & Results\\\hline
Result       & composesSolution   & Solution\\\hline
Cause        & dasAction          & Action\\\hline
Action       & belongsToScope     & Scope\\\hline
Cause        & hasTheme           & Theme\\\hline
Cause        & proposesSolution   & Solution\\\hline
Problem      & generatesCause     & Cause\\\hline
  \end{tabular}
  \label{ospRestr}
\end{table}

Also, in \vcps, existential restrictions were  written as ``min 1''. 
These were changed to the standard ``some'' existential restriction.
